---
title: "Global COVID-19 Trends: Cases, Vaccinations, and ICU Strain"
author: "Li Zhu, Abdulrahman Osilan, Mohammed Badar Al Mamari"
execute:
  echo: false
format:
  pdf:
    fig-width: 8
    fig-height: 4
    fig-cap-location: top   
    tbl-cap-location: top   
    message: false          
    warning: false          
    error: false           
    fig-pos: 'H' 
    toc: true
    code-fold: true
    theme: cosmo
editor: visual
bibliography: references.bib
csl: apa.csl
---
```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(knitr)
library(tidyr)
library(kableExtra)
library(lubridate)
library(tidyverse)
library(ggplot2)
library(scales)
library(tools)
library(viridis)  # 加载色盲友好调色板包
```

# Introduction

This project analyzes the impact of COVID-19 across major countries between 2020 and 2024. The following research questions guide our study:

1.  How did daily case counts change across countries over time?
2.  What is the relationship between vaccination and death rates in the US?
3.  How did ICU capacity respond to COVID-19 waves globally?

We use trusted global datasets, clean and transform them, and present insights using statistical summaries and visualizations.



# Data Overview

We utilize publicly available datasets from: - **Johns Hopkins University (JHU)**: Confirmed global case time series. - **COVID-19 Data Hub**: U.S.-focused metrics on deaths and vaccinations. - **Our World in Data (OWID)**: Global ICU and mortality metrics.

The OWID dataset provides comprehensive metrics [@owid].

Johns Hopkins data is widely used for case tracking [@jhu].

The COVID-19 Data Hub R package supports granular filtering [@covid19hub].

All data were preprocessed using R scripts and stored in `tidied_data/`.

# 1. COVID-19 Case Trends by Country

```{r}
#| label: tbl-summary_of_monthly_covid19_cases_by_country
#| tbl-cap: "Summary of Monthly COVID-19 Cases by Country"
jh_cases <- read_csv("tidied_data/jh_confirmed_cases_country_level.csv")

case_summary <- jh_cases %>%
  group_by(`Country/Region`) %>%
  summarise(
    `Total Cases` = sum(new_cases, na.rm = TRUE),
    `Peak Increase` = max(new_cases, na.rm = TRUE),
    `Peak Date` = year_month[which.max(new_cases)]
  ) %>%
  mutate(
    `Total Cases` = comma(`Total Cases`),
    `Peak Increase` = comma(`Peak Increase`)
  ) %>%
  arrange(desc(`Total Cases`))

kable(case_summary, 
      caption = "Summary of Total and Peak Monthly COVID-19 Cases by Country",
      align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = "scale_down",
                full_width = FALSE)
```

The table @tbl-summary_of_monthly_covid19_cases_by_country summarizes total COVID-19 case five countries as well as the peak increase on a specific date. The United States recorded the highest number of total cases—over **103.8 million**, and a peak at **January 1st 2022** with over **20.3 mllion cases**. The United Kingdom follows with a significantly lower peak, though it also peaked on the same date. All five countries observed their peak case counts at the start of **2022**, suggesting a global surge likely related to the Omicron wave. Singapore and Canada had comparatively lower case totals, Their relatively low peak increase of cases potentially reflected a more thorough safety and quarantine policy.

# 2. US Vaccination and Death Relationship

```{r}
#| label: tbl-us_covid19_vaccination_and_death_summary
#| tbl-cap: "US COVID-19 Vaccination and Death Summary"
# Load the data
us_data <- read_csv("tidied_data/covid_datahub_USA_filtered.csv")

# Compute summary statistics using daily changes
summary_stats_us <- us_data %>%
  summarise(
    Total_Deaths = sum(deaths_change, na.rm = TRUE),
    Total_Vaccinated = sum(people_vaccinated_change, na.rm = TRUE),
    Total_Fully_Vaccinated = sum(people_fully_vaccinated_change, na.rm = TRUE),
    Mean_Daily_Deaths = mean(deaths_change, na.rm = TRUE),
    Median_Daily_Deaths = median(deaths_change, na.rm = TRUE),
    Mean_Daily_Vaccinated = mean(people_vaccinated_change, na.rm = TRUE),
    Median_Daily_Vaccinated = median(people_vaccinated_change, na.rm = TRUE),
    Mean_Daily_Fully_Vaccinated = mean(people_fully_vaccinated_change, na.rm = TRUE),
    Median_Daily_Fully_Vaccinated = median(people_fully_vaccinated_change, na.rm = TRUE)
  )

# Correlation based on daily changes
correlation <- cor(us_data$people_fully_vaccinated_change, us_data$deaths_change, use = "complete.obs")
summary_stats_us$Correlation_FullyVaccinated_Deaths <- round(correlation, 3)

# Convert to long format and clean labels
summary_stats_us_long <- summary_stats_us %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") %>%
  mutate(
    Metric = gsub("_", " ", Metric),
    Metric = gsub("(?<=[a-z])(?=[A-Z])", " ", Metric, perl = TRUE),
    Metric = toTitleCase(tolower(Metric)),
    Value = format(round(as.numeric(Value)), big.mark = ",")
  )

# Render formatted table
kable(summary_stats_us_long,
      caption = "US COVID-19 Vaccination and Death Summary (Using Daily Changes)",
      align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                latex_options = "scale_down",
                full_width = FALSE)
```

This table @tbl-us_covid19_vaccination_and_death_summary summarizes U.S. COVID-19 deaths and vaccination statistics based on daily changes. The total number of deaths exceeded **836,000**, while over **269 million individuals** received at least one dose of the vaccine, and more than **230 million were fully vaccinated**. On average, there were approximately **1,006 deaths** per day, with a **median of 608**, reflecting variability in daily fatality counts. Mean daily vaccination figures reached around **324,624**, and daily full vaccinations averaged **277,251**. The correlation between daily full vaccinations and daily deaths was approximately at **0.16**, suggesting a **weak positive linear association** — This pattern is likely due to overlapping trends over time rather than a direct cause-and-effect relationship. To better understand how these factors are connected, further analysis using time series or lagged data would be needed.

# 3. ICU Response by Country

```{r}
#| label: tbl-icu_response_to_covid19_waves_by_country
#| tbl-cap: "ICU Response to COVID-19 Waves by Country"
icu_data <- read_csv("tidied_data/owid_covid_data_filtered_final.csv")

icu_summary <- icu_data %>%
  filter(!(country %in% c("China", "Singapore"))) %>% 
  arrange(country, date) %>%
  group_by(country) %>%
  mutate(
    prev_icu = lag(icu_patients),
    icu_increase = ifelse(!is.na(prev_icu) & icu_patients > prev_icu, icu_patients - prev_icu, 0)
  ) %>%
  summarise(
    Country = first(country),
    `Total ICU Patients` = format(sum(icu_increase, na.rm = TRUE), big.mark = ",", scientific = FALSE),
    `Max ICU Patients` = format(max(icu_patients, na.rm = TRUE), big.mark = ",", scientific = FALSE),
    `Peak Date` = date[which.max(icu_patients)] 
  ) %>%
  select(Country, `Total ICU Patients`, `Max ICU Patients`,`Peak Date`) %>%
  arrange(desc(`Max ICU Patients`))

# ICU Response to COVID-19 Waves by Country
kable(icu_summary, caption = "ICU Response to COVID-19 Waves by Country",
      align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

This table @tbl-icu_response_to_covid19_waves_by_country shows the overall burden on ICU systems across five countries. The U.S. again leads with over **103,680 total ICU patients**, peaking at **28,891 patients** on **January 12th 2021**. The United Kingdom and Canada followed with considerably lower ICU loads, suggesting either a more stable hospitalization trend or better capacity management.

# Visualization

## Yearly COVID-19 Cases Across Countries

```{r}
#| label: fig-yearly_covid19_case_counts
#| fig-cap: "Yearly COVID-19 Case Counts Over Time (2020–2024)"
#| fig-width: 8
#| fig-height: 6

covidTidy <- read_csv("tidied_data/owid_covid_data_filtered_final.csv") %>%
  group_by(country) %>%
  mutate(
    date = as.Date(date),
    timepoint = row_number(),
    new_cases = case_when(
      timepoint == 1 ~ 0,
      .default =  total_cases - lag(total_cases)
    )
  ) %>%
  select(-timepoint)

ggplot(covidTidy, aes(x = date, y = new_cases, color = country)) +
  geom_line(size = 0.5) +
  scale_color_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    x = "Year",
    y = "Yearly Cases",
    color = "Country/Region",
  ) +
  scale_y_continuous(labels = scales::percent_format(
    scale = 0.001,
    suffix = "k"
  )) +
  theme_minimal() +
  theme(
    legend.position = "top"
  ) +
  facet_wrap(~ country, scales = "free_y", ncol = 2)
```

As shown in @fig-yearly_covid19_case_counts. This multi-panel line chart visualizes daily COVID-19 case trends from 2020 to 2024 across five countries: Canada, China, Singapore, the United Kingdom, and the United States. The United States exhibits the most dramatic surge, peaking in early 2022 with daily counts exceeding **4 million cases**, followed by more sustained waves. Canada and the UK also experienced intense waves centered around early 2021 and 2022. In contrast, China's data shows a singular, sharp peak, likely associated with a delayed but explosive wave post-lockdown. Singapore’s trend reveals frequent, smaller spikes, suggesting recurring outbreaks with faster containment. These synchronized varied patterns reflect how national policies and variant emergence shaped case trajectories globally.

## Vaccination vs. Death Rates in the U.S.

```{r}
#| label: fig-vaccination_rate_vs_death_rate_intheus
#| fig-cap: "Vaccination Rate vs. Death Rate in the U.S. (2021–2023)"
covidUSA <- read_csv("tidied_data/covid_datahub_USA_filtered.csv") %>%
  mutate(
    date = as.Date(date),
    pct_fully_vaccinated = (people_fully_vaccinated / 335000000) * 100,
    daily_deaths = deaths - lag(deaths),
    death_rate_per_100k = (daily_deaths / 335000000) * 100000,
    year_date = as.factor(year(date))
  ) %>%
  filter(pct_fully_vaccinated > 0, daily_deaths > 0)

ggplot(covidUSA, aes(x = pct_fully_vaccinated, y = death_rate_per_100k, color = year_date)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_color_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  geom_smooth(method = "loess", se = FALSE, color = "#0072B2") +
  labs(
    x = "% Fully Vaccinated",
    y = "Deaths per 100,000 People",
    color = "Year"
  ) +
  theme_bw() +
  theme(
      legend.position = "top"
    )

```

As shown in @fig-vaccination_rate_vs_death_rate_intheus. This scatterplot shows the relationship between full vaccination rates and COVID-19 deaths per 100,000 people in the United States, colored by year. The smoothed curve reveals a clear inverse relationship during 2020–2021, where increased vaccination coverage is associated with a substantial drop in mortality. However, this trend appears to have reversed in 2022, with higher vaccination rates not necessarily leading to lower death rates. This suggests that other factors, such as the emergence of new variants or changes in public health policies, may be influencing COVID-19 outcomes. By late 2023, the fully vaccinated population increases significantly, reaching nearly **70% in 2024**. During this period, the death rate becomes more controlled, suggesting either improved vaccine effectiveness or a reduction in the severity of circulating variants.

## ICU Occupancy Over Time

```{r}
#| label: fig-monthly_icu_occupancy_during_covid19
#| fig-cap: "Monthly ICU Occupancy During COVID-19"
covid_ICU <- read_csv("tidied_data/owid_covid_data_filtered_final.csv") %>%
  mutate(
    date = as.Date(date),
    date_ym =  as.Date(format(date, "%Y-%m-01"))
  ) %>%
  filter(country %in% c("United States", "United Kingdom", "Canada")) %>%
  group_by(country, date_ym) %>%
  summarise(icu_patients = sum(icu_patients, na.rm = TRUE), .groups = "drop") %>%
  mutate(icu_patients = icu_patients / 1000)

ggplot(covid_ICU, aes(x = date_ym, y = icu_patients, fill = country)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  annotate("rect", xmin = as.Date("2021-06-01"), xmax = as.Date("2021-11-01"), ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "#009E73") +
  annotate("rect", xmin = as.Date("2021-11-01"), xmax = as.Date("2022-02-01"), ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "#999999") +
  annotate("text", x = as.Date("2021-08-01"), y = 900, label = "Delta", size = 3.5) +
  annotate("text", x = as.Date("2022-01-01"), y = 900, label = "Omicron", size = 3.5) +
  labs(
    x = "Month",
    y = "ICU Patients (in thousands)",
    fill = "Country"
  ) +
  scale_x_date(
    breaks = seq(as.Date("2020-01-01"), as.Date("2024-08-01"), by = "3 months"),
    date_labels = "%Y-%m"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
    )
```

As shown in @fig-monthly_icu_occupancy_during_covid19. This grouped bar chart illustrates monthly ICU occupancy (in thousands) across the U.S., U.K., and Canada **from 2020 through 2024**. The United States shows dominant ICU demand throughout the period, particularly during **end of 2020-start of 2021**, the **Delta** and **Omicron waves**, which are shaded for reference. ICU occupancy in the U.S. alone peaked at over **800,000 patients**. The **Delta period (mid-2021)** aligns with a sharp rise in ICU patients, peaking above 800,000(Three countries). The U.K. and Canada exhibit far lower but synchronized ICU trends. Notably, ICU burden in all three countries dropped significantly after the Omicron wave, possibly due to improved treatments, increased vaccine protection, and lower hospitalization rates in later variants. This figure emphasizes the pressure variant surges placed on critical care infrastructure.

# Conclusion

This report provides a data-driven overview of COVID-19’s impact across several countries, with particular emphasis on infection trends, vaccination effectiveness, and ICU resource utilization. The analysis reveals that the United States and the United Kingdom experienced the highest case volumes and ICU burdens, both peaking in January 2022—likely driven by the Omicron variant. Despite stark differences in healthcare infrastructure and containment strategies, all five countries displayed synchronous peak patterns, underscoring the global nature of COVID-19 surges.

Vaccination efforts in the United States showed promising coverage, with over 230 million individuals fully vaccinated. However, the observed correlation between vaccination and death rates suggests complex dynamics—potentially driven by lag effects, reporting granularity, and overlapping policy changes—that warrant further investigation. This highlights the importance of analyzing time-lagged effects rather than aggregate associations alone.

ICU data further underscore the disparity in health system strain. While the U.S. and U.K. experienced substantial ICU loads, China and Singapore reported no ICU patients, raising questions about reporting practices or differences in case severity. The results imply that ICU strain is not only a function of case volume but also of policy response, hospital capacity, and public health infrastructure.

Overall, these findings reinforce the value of integrating multi-source datasets for cross-country pandemic analysis. For future studies, incorporating population-adjusted metrics (e.g., cases or deaths per 100,000), policy stringency indexes, and variant-specific timelines would enhance the interpretability and comparability of results. Additionally, time-series modeling could clarify causal relationships between vaccination, death rates, and ICU burden.

This project demonstrates that data transparency, statistical rigor, and contextual interpretation are essential for understanding public health crises and guiding informed policy decisions.

# References

::: {#refs}
:::

# Appendix: Full Code
```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE} 
```